---
title: What's The Difference? An Analysis of Hours and Productivity of Germany and
  Poland
author: "John Brunolli"
date: "May 6, 2019"
fontsize: 12pt
geometry: margin = 1in
output: pdf_document 
---

\section{Introduction}

Have you ever wondered what life would be like with a shorter work week? I thought about 

that too when I was a full-time worker. But I look at European countries, I overall notice 

they work less hours than American workers and that got me thinking about what the 

economic outcomes may happen if the United States goes along with this trend. After 

researching some popular articles and some academic papers behind labor economics, 

they generally show that European countries have had their full-time hours follow a 

downward trend. But however, testing every European country would take a lot of time 

since there are plenty nations in the European continent. Therefore, conducting a simple 

standard OLS regression may not be the most optimal action to partake in for this task, 

therefore I am going to use difference in difference approach. Now the natural question is 

which European countries shall I use for the difference in difference estimation? After  

long and careful considerations, I have decided to test Germany and Poland.  

I wish to pursue this idea because this is an interesting proposition from an economics point 

of view. That is if I am producing at the same rate but at a shorter work schedule then would 

it make sense for countries to follow this natural trend. This trend has started in Germany 

after major unions, such as the IG Metal union in Germany, won major bargaining agreements 

with firms in regard to negotiations of working hours. Their main purpose behind these hours 

is to give workers a chance to enjoy life outside from work. That does make sense because 

having time with family is important, time away from work can help make some workers 

happy and makes working conditions a little more bearable. In 2008 that's when other 

German industries also decided to adopt this idea too and that's why we observe Germany 

work at around 35 hours a week. All these are good and reasonable, then comes the natural 

question what will happen to productivity of the entire German continent? Also, what will 

happen to the overall national wages of the workers? That is what I am attempting in this 

model, but why did I choose Poland as a comparison?   

Then arises the natural question, why Germany and Poland? Why not France and Germany, 

Germany and the Czech Republic, why those two. I selected these countries based on a 

couple of different criteria. First both countries are right next to each other. That 

argument can be said with Belgium, France, the Czech Republic, but I'll elaborate further 

towards my criteria. But if I chose Italy and Germany I may get some concerns arising 

since there is a good amount of distance between these countries, but I remember writing 

a paper on Italian culture and a good amount of Italians sometimes distance themselves 

from other European countries because they feel they have a different culture. That is 

very subjective, but that's a flaw I saw with a country like Italy, and certainly 

climate and policies sure impacted my decision too. Now the main reason why I chose these 

two countries is based on the trends I observe from these two unique European countries.

Primarily since according to the trends within the data and popular sources, Germany is 

exhibiting a downward trend in working hours and a lot of Germans work 35-hour days on 

average. While Poland has one of the longer workdays within the European continent. Both 

countries have similar work laws on the books, but the only difference is that Germany is 

on a downward trend in their working hours, while Poland still maintains some of the 

longest working hours in Europe. This can be illustrated with both of these graphs. 

```{r, include=FALSE}
library(readxl)

DataFinal_1 <- read_excel("C:/Users/rbrun/Downloads/DataFinal_1.xlsx", 
                          col_types = c("text", "text", "numeric", 
                                        "numeric", "numeric", "numeric", 
                                        "numeric", "numeric", "numeric", 
                                        "numeric", "numeric", "numeric", 
                                        "numeric", "numeric", "numeric", 
                                        "numeric", "numeric", "numeric")) 

Poland <- read_excel("C:/Users/rbrun/Downloads/Poland.xlsx",
                            col_types = c("text", "text", "numeric", 
                                           "numeric", "numeric", "numeric", 
                                           "numeric", "numeric", "numeric", 
                                           "numeric", "numeric", "numeric", 
                                           "numeric", "numeric", "numeric", 
                                           "numeric", "numeric", "numeric"))

Germany <- read_excel("C:/Users/rbrun/Downloads/Germany.xlsx", 
    col_types = c("text", "text", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric"))
```

```{r, include=TRUE, echo=TRUE}
plot(Germany$Time, Germany$Hours)

plot(Germany$Time, Germany$Productivity)

plot(Germany$Time, Germany$Wage)
```

Here I notice a downward trend within the German working hours over time. With this plot 

clearly demonstrating that the trend is decreasing, with the occasional increase in the 

working hours, but pretty much have decreased. Notice though that the productivity has 

continued to rise for Germany even though that goes against neoclassical labor production 

theory I should observe production to decrease. The trend is a little awkward if you ask me. 

Also, here wage is continuing to increase over time. Even when the trend for hours worked in 

a work week is decreasing. That makes my question more interesting because gaining 

knowledge on this trend will only be an effective way of how to carry out and manage a 

successful economy. 

```{r, include=TRUE, echo=TRUE}
Poland_1=Poland[c(-1, -2, -3, -4, -5,-6),]
#View(Poland)
plot(Poland_1$Time, Poland_1$Hours)
plot(Poland_1$Time, Poland_1$Productivity)
plot(Poland_1$Time, Poland_1$Wage)
```

Here what I notice for Poland is that there working hours for the most part remain consistent 

up until 2008 and then there was a decrease in working hours, probably due to the economic 

recession in 2008, but their hours are still higher than Germany's. Also, their wages and

production have continually increased and that does coincide with labor and macroeconomic 

theory regarding working hours and productivity and wage. Based on these graphs and trends

within both economies, I am curious to see what the overall effects of the German's economy's 

trend towards shorter working hours are have on their productivity and wage. I am going to 

implement a difference and difference estimation in order to effectively test this by comparing

the German and Polish economies. My analysis should hopefully gain insight into whether 

neoclassical production theory and macroeconomic principles will hold and therefore, I will 

expect the German economy to see a decrease in their wages and their production. 

\section{Methodology}

I gathered my data from the OECD database website and I composed into three datasets.

I composed the data into three datasets because first I needed to make references to 

what are the trends within the data, so I collected each of the two countries 

individually first. Then I collected a third dataset with each of these two countries

together in order to conduct the DID estimation. I couldn't combine the two datasets 

in R, therefore I had to collect another dataset, but I kept every variable constant.

I personally think that a standard OLS regression for both the German and Polish economy 

may explain a correlation, but will that be efficient? Therefore, the following equations:

$$
Production= \beta_0 + \beta_1*Hours + \epsilon
$$

and

$$
Wage=\beta_0 + \beta_1*Hours + \epsilon
$$

These may provide good insight for correlation, but that's not what's at stake here. Here I 

am trying to observe some potential causality between wage and production and hours. 

After careful examination of the data and the methods I can implement, I have chosen to go 

with the difference in difference approach since that makes the most sense to implement in 

this case considering that having an instrumental variable to may explain causation for 

Germany, butfails to explain Poland. Also, an RDD here I think would be too difficult to 

implement considering the nature of the data. Therefore, the formal equations I will implement 

are: 

$$
Production= \beta_0 + \beta_1*TD + \beta_2*CD + \beta_3*DID + \epsilon
$$

and 

$$
Wage= \beta_0 + \beta_1*TD + \beta_2*CD + \beta_3*DID + \epsilon
$$

Where TD is my time variable where I generated a dummy variable where 1 equals the year 

greater than or equal to 2008, 0 if otherwise. CD is the dummy variable I generated where 1 

equals Germany, 0 if Poland. The DID is my difference between when the 35-hour work week 

trend started for Germany and the time. This difference in difference estimation equation 

will provide a better understanding of the causality between implementing a 35-hour work 

week and what the impacts the policy has on productivityand wages. That was I will observe 

the treatment effect of this policy in regards to what happened to Germany compared to 

Poland. Now of course there was issues with the data, mainly incomplete entries within the 

variables of measure, therefore I had to clean up the data in order to actually proceed 

further with the tests and analysis. After, that issue was resolved, then I generated my 

difference in difference estimation and just to double check on my adjusted R^2 and my R^2 

I ran a Breusch-Pagan test to determine if heteroskedasticity is present and I ran a 

Shapiro-Wilks test to test to see if the model is normal.

\section{Results}

In order to run my analysis, I had to clean up some data because there were some missing 

variables for Poland. That can be attributed to the cold war since Poland was under Soviet 

control until 1989. After the data was cleaned up, I finally implemented my difference in 

difference estimation in regards to production. 

```{r, include=TRUE}
GerPol=DataFinal_1[c(-1, -2, -3, -4, -5, -6, -7, -8, -9, -10, -11),]
GerPol$TD=ifelse(GerPol$Time >= 2008, 1, 0)
#View(GerPol)
GerPol$CD=ifelse(GerPol$Country == "Germany", 1, 0)
GerPol$did=GerPol$TD * GerPol$CD
```

```{r, include=TRUE, echo=TRUE}
modeldid=lm(Productivity~TD + CD + did, data = GerPol)
summary(modeldid)
```

These numbers are what the neoclassical production theory tells us and that's the less hours 

worked then there will be less production. There could also be factors at play here that could 

explain the production loss. For example, take the time, simply the time interprets that on 

average 2008 and beyond was 31% more productive than Poland in between 2008 to 2019. 

That gets to my main assertion because there was a global recession, hence that could explain

why we see the drop-in production. Also, Germany on average produced 17% more than 

Poland in general. But overall the main interpretation from this model is that when Germany 

started to follow the trend and worked on average for hours their production decreased by 

24% compared to Poland who didn't follow suit. But overall the treatment effect shows that 

when Germany lowered it's work week hours to 35, it experienced a 24% loss of production. 

But there's another aspect I need to examine, how will the 35-hour work week affect the 

wages of the workers. Using my DID on Wages I found the following: 

```{r, include=TRUE, echo=TRUE}
modeldidw=lm(Wage~TD + CD + did, data = GerPol)
summary(modeldidw)
```

These results following what I expect from my experiences with macroeconomic theory. As I 

observe on average the 35-hour work week has decreased the income of Germany by 12,834 

Euros. Compared to Poland who didn't implement the 35-hour work week, this can be 

concerning considering that no one wants to see their incomes decrease especially by tens 

of thousands of dollars. What can be impacting this? Looking back at the time frame, there 

was a global economic recession back in 2008 and that could be impacting these results. 

That would not be shocking to observe the impact of the recession, especially since the 

recession did cause a lot of layoffs of workers globally, therefore wages would naturally 

decrease. All in all, I am not surprised with the results and is consistent with standard 

macroeconomic theory. But I will check both of these models for normality and 

heteroskedastictiy just to check the assurance of the DID models I generated.

```{r, include=TRUE}
library(lmtest)
```

```{r, include=TRUE, echo=TRUE}
bptest(modeldid)
resdid=modeldid$residuals
shapiro.test(resdid)
```

Here I used a Breusch-Pagan test to test for heteroskedasticity present within my models. 

I used the Breusch-Pagan test because I am more confident in that test than I am with the

White's test or another way to measure. Also, I am testing for normality using a 

Shapiro-Wilk normality test. After conducting these tests on my productivity difference 

and difference estimation I can conclude that there is heteroskedasticity present in my 

model and based on the Shapiro-Wilk test, I can conclude that the model is normal. 

Heteroskedasticity can be a problem or a concern, but here I am not concerned because 

considering the nature of the data being randomly collected, I think that this is not a 

problem. Also, the fact that my R^2 and adj R^2 are relatively high, I am confident that 

this model is a good step in the right direction. 

```{r, include=TRUE, echo=TRUE}
bptest(modeldidw)
resdidw=modeldidw$residuals
shapiro.test(resdidw)
```

So, can heteroskedasticity affect the validity of my model? Here I don't think so because 

given the nature of the data I am analyzing, wages can vary within thousands of dollars, 

the heteroskedasticity present may not be a problem. That's because the data by nature is 

already going to have vastly different ranges of values that the wage can take. So overall, 

the wage DID estimation is still valid, even though heteroskedasticity is present, because 

of the nature of the data and the satisfactory levels of R^2 and adj R^2. 


\section{Discussion}

As I stated earlier my results generated from my DID regression are not at all surprising. 

These are consistent with what neoclassical production theory and macroeconomic theory 

predicts what would happen. But there are some concerns the results give? Mainly because 

take the wage, the wage of a German citizen is on average higher than the average Polish 

citizen. Then it is natural to ask, why does Germany still have more income than Poland at 

least with raw data? It could tie back to again West Germany was not part of Soviet control, 

East Germany and Poland were under Soviet control, therefore, it may be possible that 

Germany when the country was recombined, benefitted greatly from the reuniting of the 

country and therefore minimizing the overall effect of being under Soviet controlled, 

while Poland doesn't have that luxury. Also, there could be other factors I am not considering 

in my model such as Germany's strong education program, technological advantages, and 

political stability that may affect this that I am not able to measure. The recession as 

I elaborated multiple times in the results section could be greatly impacting the results, 

considering I am controlling within 2008 to 2010, during the height of the global recession 

can significantly impact these results. Maybe you'd think Poland is less protected than 

Germany, but again the results are demonstrating trends within the 35-hour work week trend. 

Then the natural question is asked about can we trust the models? Certainly, looking at 

the R^2 and the adj R^2 both models appear to be doing a pretty good job of explaining the 

noise within the data. The standard errors for the production models are very small, 

therefore I have confidence in that model. However, in the wage DID, there were high 

standard errors with the treatment effect being at 2,649. That is concerning from a raw 

numbers point of view but considering that wages are within tens of thousands of dollars, 

is it necessarily a bad thing with this high standard error, given the data I am working 

with? I mean for sake of argument, I will assume that the on average the 35-hour work week 

decreases in earnings by 10,000, I predicted roughly 12,000. That is not that far off 

because the data is wages of both countries. I am fairly confident with this model, I am not 

that concerned about heteroskedasticity being present considering that I do have 

favorable R^2's and adj R^2's in both models and the fact that each of these variables 

are statistically significant is reassuring that the treatment effect has something to 

do with the taking on of the 35-hour work week. Overall, I am pleased with this model and 

is a good starting point in figuring out the overall effects of the adoption of a 35-hour 

work week. 

\section{Conclusion}

My models I generated have overall determined what I expected from a labor and 

macroeconomic point of view. The model showed that when Germany have implemented the 

trend in their country of a 35-hour work week, I observed that production and wages 

fell. There were some surprising details there to as well, and perhaps contradicting 

such as on average just by being in Germany and being a Polish citizen might yield more 

money than German's are what the resultsare telling me. But the big question is did this 

model solved the riddle. Of course not, I think this is a good stepping stone, however 

there are limitations to this model. One is the fact that the model cannot account for

the effects of the recession. I am confident that the recession is affecting my results 

without a doubt, but is that affect stronger than the effect of the 35-hour work week?

Also, are there other factors that I am not accounting for such as Germany's political 

stability vs Poland's instability. How impactful is the German education system on 

the wage and productivity considering Germany has strong scores in educational

measurements, while Poland is not there? With this model will Germany stop this trend 

towards a 35-hour work week. I doubt it since that the country has strong labor unions 

and one of the most powerful labor unions won a 28-hour work week for its workers now 

it might be inevitable that Germany will adjust its work weeks. All in all, this project 

has been interesting from many elements, but I am pleased with the model and the 

implications from the following implementation of the 35-hour work week.

\newpage

\section{Works Cited}

Burda, Michael C., et al. "Cyclical Variation in Labor Hours and Productivity Using 

the ATUS." The American Economic Review, vol. 103, no. 3, 2013, pp. 99-104. JSTOR, 

www.jstor.org/stable/23469710.

Petroff , Alanna Petroff. "German Workers Win Right to 28-Hour Week." CNNMoney, 

Cable News Network, 7 Feb.2018,

money.cnn.com/2018/02/07/news/economy/germany-28-hour-work-week/index.html.

"Features of German Labor and Employment Law." WilmerHale, 8 Oct. 2003, 

www.wilmerhale.com/en/insights/publications/features-of-german-labor-and-employment-law-october-8-2003

Germany Trade and Invest GmbH. "Terms of Employment." GTAI, Germany Trade & Invest, 2019, 

www.gtai.de/GTAI/Navigation/EN/Invest/Investment-guide/Employees-and-social-security/terms-of-employment.html.

"Labour Regulations." Www.paih.gov.pl, Polish Development Fund, 2019, 

www.paih.gov.pl/polish_law/labour_regulations.

Oecd. Economic Outlook No 104 - November 2018., OECD, 2018, 

stats.oecd.org/viewhtml.aspx?datasetcode=EO104_INTERNET&lang=en#.

OpenStax. "Principles of Economics: Chapter 4." Principles of Economics, Pressbooks, 2019, 

opentextbc.ca/principlesofeconomics/chapter/4-1-demand-and-supply-at-work-in-labor-markets/#Table_04_0

2.

Sarva, Amol. "Why Germans Work Fewer Hours But Produce More: A Study In Culture." 

HuffPost, 7 Dec. 2017, www.huffpost.com/entry/why-germans-work-fewer-ho_b_6172262.
